Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-11-05 paper/abstract.tex
changes to main_figure and paper

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs to reveal
their rotational evolution and demonstrate that kinematics have potential to
be a useful tool for calibrating gyrochronology.
At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
to the Praesepe cluster, accurately predicts the relative ages of stars.
At these young ages, lower-mass stars spin more slowly than higher-mass stars
because their stronger magnetic fields lead to more efficient angular momentum
loss.
However, at old ages we find that late G and early K dwarfs rotate {\it
faster} than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results, based on field star rotation periods, align with recent
findings from the rotation periods of stars in middle-aged open clusters and
theoretical models that vary the rate of surface-to-core angular momentum
transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
proxy, to reveal their rotational evolution and demonstrate that kinematics
could be a useful tool for calibrating gyrochronology.
At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
to the Praesepe cluster, accurately predicts the relative ages of stars.
At these young ages, lower-mass stars spin more slowly than higher-mass stars
because their stronger magnetic fields lead to more efficient angular momentum
loss.
However, at old ages we find that late G and early K dwarfs rotate {\it
faster} than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results, based on field star rotation periods, align with recent
findings from the rotation periods of stars in middle-aged open clusters and
theoretical models that vary the rate of surface-to-core angular momentum
transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

### 2019-11-05 paper/discussion.tex
changes to main_figure and paper

The results presented above indicate that stars of spectral type ranging from
late G to late K follow a braking law that changes over time.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.

Our results indicate that, at young ages, stellar rotation period decreases
with mass.
This is probably because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfven radii and therefore experience greater
angular momentum loss rate.
According to the \citet{spada2019} models, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation periods appear constant with mass, and at late
ages rotation period increases with mass for GK dwarfs.
The explanation for this phenomenon, according to the \citet{spada2019}
models, is that lower-mass stars are still braking more efficiently at these
intermediate and old ages but their cores are more tightly coupled with their
envelopes, so angular momentum resurfaces and prevents the stellar envelopes
from spinning-down rapidly.

The origin of the rotation period gap, first identified by
\citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
\ref{fig:dispersion_period_teff} still remains a mystery.
This gap can be seen as an under-density of points between the 0.7-1.0 and
1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
of constant gyrochronal age of around 1.1 Gyr \citep[according to the
gyrochronology relation of][]{angus2019}, as shown in figure
\ref{fig:dispersion_period_teff}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018} and a change in magnetic field structure causing a brief period
where rotational variability is reduced and rotation periods cannot be
measured \citep{reinhold2019}.
Our results indicate there might be a slight difference in rotational
evolution below and above the gap.
Stars below the gap appear to be in the regime where wind-braking is dominant
(rotation period decreases with increasing mass at a given age) and stars
above the gap are in a regime where rotational coupling is dominant (rotation
period increases with increasing mass at a given age).
This can be seen in figures \ref{fig:age_cut} and \ref{fig:vplot}.
For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
velocity dispersion increases with \teff.
In figure \ref{fig:age_cut}, velocity dispersion within a given period range
appears to {\it decrease} with decreasing temperature below the gap.
The opposite appears to be true above the gap.
This suggests that perhaps stars transition from a wind braking regime to a
rotational coupling regime as they pass across the gap.
However, it is unclear how this process could {\it create} the gap.
A gap could be created if stars rapidly lose angular momentum over a short
period of time, leading to a dearth of stars in the gap.
However, a sudden redistribution of angular momentum from the core to the
surface, should cause stellar rotation periods to {\it decrease}, not
increase.
In addition, this theory is not supported by observations of the NGC 6811
cluster whose K dwarf members have already undergone stalled braking below and
before they reach the gap.
The gap may be associated with a wind braking-rotational coupling transition,
however our results only provide weak evidence for this, so we can only
speculate until more data become available.

>>>>

The results presented above indicate that stars of spectral type ranging from
late G to late K follow a braking law that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.

The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages, stellar rotation period decreases with mass.
This is probably because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfven radii and therefore experience greater
angular momentum loss rate.
According to the \citet{spada2019} models, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for GK dwarfs.
The explanation for this, according to the \citet{spada2019} models, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two interior layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly.

% The origin of the rotation period gap, first identified by
% \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
% \ref{fig:dispersion_period_teff} still remains a mystery.
% This gap can be seen as an under-density of points between the 0.7-1.0 and
% 1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
% of constant gyrochronal age of around 1.1 Gyr \citep[according to the
% gyrochronology relation of][]{angus2019}, as shown in figure
% \ref{fig:dispersion_period_teff}.
% Several explanations for the gap's origin have been proposed, including a
% discontinuous star formation history \citep{mcquillan2013, davenport2017,
% davenport2018} and a change in magnetic field structure causing a brief period
% where rotational variability is reduced and rotation periods cannot be
% measured \citep{reinhold2019}.
% Our results indicate there might be a slight difference in rotational
% evolution below and above the gap.
% Stars below the gap appear to be in the regime where wind-braking is dominant
% (rotation period decreases with increasing mass at a given age) and stars
% above the gap are in a regime where rotational coupling is dominant (rotation
% period increases with increasing mass at a given age).
% This can be seen in figures \ref{fig:age_cut} and \ref{fig:vplot}.
% For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
% \ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
% temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
% velocity dispersion increases with \teff.
% In figure \ref{fig:age_cut}, velocity dispersion within a given period range
% appears to {\it decrease} with decreasing temperature below the gap.
% The opposite appears to be true above the gap.
% This suggests that perhaps stars transition from a wind braking regime to a
% rotational coupling regime as they pass across the gap.
% However, it is unclear how this process could {\it create} the gap.
% A gap could be created if stars rapidly lose angular momentum over a short
% period of time, leading to a dearth of stars in the gap.
% However, a sudden redistribution of angular momentum from the core to the
% surface, should cause stellar rotation periods to {\it decrease}, not
% increase.
% In addition, this theory is not supported by observations of the NGC 6811
% cluster whose K dwarf members have already undergone stalled braking below and
% before they reach the gap.
% The gap may be associated with a wind braking-rotational coupling transition,
% however our results only provide weak evidence for this, so we can only
% speculate until more data become available.

### 2019-11-05 paper/intro.tex
changes to main_figure and paper

It is well established that magnetized stellar winds cause the rotation
periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
calibrated, the relationship between period, mass and age can be used to
age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
meibom2011, meibom2015}.
% However, the exact nature of rotational evolution is still being actively
% studied.
% Once the relationship between rotation period, mass and age is well
% calibrated, rotation periods can be used to precisely and accurately date
% stars via gyrochronology \citep[\eg][]{kawaler1989, pinsonneault1989,
% barnes2003, barnes2007, barnes2010, meibom2011, meibom2015}.
% It is also well established that stars of the same age but different masses
However, new photometric rotation periods measured from \kepler\ light curves
reveal that rotational evolution is more complicated than previously thought.
For example, the M dwarfs in the $\sim$ 700 Myr Praesepe cluster spin more
slowly than the G dwarfs -- in theory because lower-mass stars have deeper
convections zones which generate stronger magnetic fields and more efficient
magnetic braking \citep{schatzman1962, weber1967}.
However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
same} rate as early-K dwarfs \citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period-color relation \citep{spada2019}.
These models suggest that mass and age-dependent coupling between the cores
and envelopes of stars has a significant impact on their surface rotation
rates.

Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
the observations \citep{vansaders2016, vansaders2018}.
% revealing that rotational
% evolution is clearly a more complicated process than once thought.
% However, an underlying assumption behind many empirical gyrochronology
% relations is that the relationships between rotation period and photometric
% color\footnote{As the directly observable quantity, color is often used as a
% mass proxy, and empirical gyrochronology relations are usually calibrated in
% color, rather than mass or effective temperature.},
% and rotation period and age are {\it separable}, meaning that the
% period-color relation is the same at all ages and the period-age relation is
% the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
% meibom2011, angus2015, angus2019}.
% Not all gyrochronology models follow this form, particularly those based on
% stellar structure and evolution theory or with a Rossby number, rather than
% rotation period dependence \citep[\eg][]{barnes2010, matt2012,
% vansaders2016]}.
% For example, old field stars seem rotate more rapidly than models predicted
% \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}, and a
% mass-dependent modification to the classical \prot\ $\propto t^{\frac{1}{2}}$
% spin-down law \citep{skumanich1972} is required to reproduce the data
% \citep{vansaders2016, vansaders2018}.
% In addition, recent analyses of the 1.1 Gyr open cluster, NGC 6811, provide
% further indication that rotational evolution is mass and time-dependent
% \citep{curtis2019 spada2019}.
% This cluster has a flattened relationship between rotation period and
% color: the early-K dwarfs rotate at the same rate as the late-K dwarfs.
% With the advent of large time-series photometry surveys, providing large
% numbers of precise rotation period measurements over the last few years.

The rotational evolution of stars is clearly a complicated process, and to
fully calibrate the gyrochronology relations we need reliable ages for large
numbers of field stars which span a range of ages.
% however relying on open clusters (which are typically young) and asteroseismic
% stars (which are typically G-type stars or more massive) to calibrate
% gyrochronology relations limits the mass and age coverage of the calibration
% sample.
In this paper, we use the velocity dispersions of field stars to gain insight
into their rotational evolution.
% This technique provides relative ages for groups of stars that cover a broad
% range of masses and ages, allowing us to observe the rotational evolution of
% stars in an average sense.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the galaxy with a low out-of-plane, or vertical, velocity ($W$, or \vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of stars increase over time
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
Although the cause of dynamical heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of field stars in
the Galactic thin disk to ascertain which groups are older and which younger
and draw conclusions based on the implied relative ages.
% The age-velocity dispersion reations (AVRs) are still actively being

>>>>

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and spin more slowly over time through magnetic braking
\citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are typically born with random rotation periods, ranging from 1
to 10 days, observations of young open clusters reveal that their rotation
periods converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009}.
After this time, it is thought that the rotation period of a star is
determined, to first order, by its color and age alone.
This is the principle behind gyrochronology -- the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
% It is well established that magnetized stellar winds cause the rotation
% periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
% weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
% calibrated, the relationship between period, mass and age can be used to
% age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
% meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} reveal that rotational evolution is more complicated
than previously thought.
For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
slowly than the G dwarfs -- in theory because lower-mass stars have deeper
convections zones which generate stronger magnetic fields and more efficient
magnetic braking.
However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
same} rate as early-K dwarfs \citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period-color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.

Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (ratio of rotation period to
convective turnover time) of around 2 \citep{vansaders2016, vansaders2018}.

The rotational evolution of stars is clearly a complicated process, and to
fully calibrate the gyrochronology relations we need reliable ages for large
numbers of field stars which span a range of ages.
% however relying on open clusters (which are typically young) and asteroseismic
% stars (which are typically G-type stars or more massive) to calibrate
% gyrochronology relations limits the mass and age coverage of the calibration
% sample.
In this paper, we use the velocity dispersions of field stars to gain insight
into their rotational evolution.
Velocity dispersion provides relative ages for groups of stars, allowing us to
observe the rotational evolution of stars over a range of masses and ages.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the galaxy with a low out-of-plane, or vertical, velocity ($W$, or \vz),
just like the star-forming molecular gas observed in the disk today
\citep[\eg][]{stark1989, stark2005, aumer2009, martig2014, aumer2016}.
On average, the vertical velocities of stars increase over time
\citep[\eg][]{nordstrom2004, holmberg2007, holmberg2009, aumer2009,
casagrande2011}.
Although the cause of dynamical heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity {\it dispersion} of a {\it group} of stars can
indicate whether, on average, that group is old or young relative to other
groups.
In this work we compare the velocity dispersions of groups of field stars in
the Galactic thin disk to ascertain which groups are older and which younger
and draw conclusions based on the implied relative ages.
% The age-velocity dispersion reations (AVRs) are still actively being

### 2019-11-05 paper/method.tex
changes to main_figure and paper

Visual binaries and subgiants were removed from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and visual binaries.
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 6th magnitude
in \gaia\ G-band.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

@@

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because only stars
with {\it detectable} rotation periods are included in it.
Since lower-mass stars stay active for longer it is likely that the oldest
stars in this sample have low masses.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD.
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
outliers before calculating the standard deviation of stars in each bin.
These high velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion
as a function of effective temperature (calculated by transforming dereddened
\gaia\ colors using equation \ref{eq:curtis}).
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
the opposite of the trend expected for mass-dependent heating, however the
slope is only inconsistent with zero at the 1.3 $\sigma$ level.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point is that we do not find evidence for mass-dependent heating
acting on stars in the \kepler\ field, meaning \sigmavb\ {\it can} be used as
an age proxy.
We also examined the {\it vertical} velocities of the 537 stars in this sample
with RV measurements.
Again, we found no evidence for mass-dependent heating: the slope of the
velocity dispersion-temperature relation was consistent with zero.

@@

>>>>

Visual binaries and subgiants were removed from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and visual binaries.
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 6th magnitude
in \gaia\ G-band.

The dwarf stars in the \mct\ sample are shown on a \gaia\ color-magnitude
diagram (CMD) in figure \ref{fig:age_gradient}.
The stars are colored by their gyrochronal age, calculated using the
\citep{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
This suggests that either old M dwarfs are missing from the \mct\ catalog, or
the \citet{angus2019} gyrochronology relation under-predicts the ages of
low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2011}, and are therefore more likely to have measurable
rotation periods at old ages, the latter scenario seems more likely.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

@@

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because only stars
with {\it detectable} rotation periods are included in it and since lower-mass
stars stay active for longer it is likely that the oldest stars in this sample
have low masses.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at the 3$\sigma$ level to remove high velocity
outliers before calculating the standard deviation of stars in each bin.
These high velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
function of effective temperature \footnote{calculated by transforming
dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
\kepler\ dwarf sample.
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
the opposite of the trend expected for mass-dependent heating, however the
slope is only inconsistent with zero at the 1.3 $\sigma$ level.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point however, is that we do not see evidence for mass-dependent
heating acting on stars in the \kepler\ field, indicating that \sigmavb\ {\it
can} be used as an age proxy.
We also examined the {\it vertical} velocities of the 537 stars in this sample
with RV measurements.
Again, we found no evidence for mass-dependent heating: the slope of the
velocity dispersion-temperature relation was consistent with zero.

@@

To explore the relationship between rotation period, \teff\ and age, we
removed high and low velocity outliers from the \mct\ sample by performing
3$\sigma$ sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, however the overall trends remained the
same.
We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
3500 K to avoid biases caused by the selection function at the faint, cool
end, and binarity or weakened braking \citep{vansaders2016} at the hot end.

% We then selected groups of stars within different gyrochronal age ranges and
% calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
% of effective temperature for each age group.
% Ages were calculated using dereddened \gaia\ \gcolor\ color, however
% throughout this paper we show rotation periods as a function of effective
% temperature, \teff, because it is easier to divide stars into bins of roughly
% equal numbers in \teff-space than in color-space.
